Analyse gender in music psychology from journal publication data

### 2025-12-12 manuscript/manuscript-apaquarto.qmd
First submission

abstract: "Academia suffers from various issues regarding equity. One such issue is gender inequality in authorship, where many studies show that women are often underrepresented. Due to authorship differences across disciplines, the goal of this study is to provide a current overview the gender distribution of authorship within music psychology. A total of 3,373 papers published between 2000 and 2025 across five core music psychology journals were analyzed with respect to author gender and authorship role (first, last, solo, and coauthor). In addition, we explored gender patterns related to citations, open-access publishing, and keywords. Overall, female authors were in the minority, accounting for 40.2% of all authors. However, the distribution of authorship roles diverged notably from this overall proportion: women were more likely to be first authors (OR = 1.41), equally likely to be solo authors (odds ratio, OR = 1.11) or coauthors (OR = 1.00), but less likely to be last authors (OR = 0.73). These patterns have become more pronounced over the past 15 years. No substantial gender differences in citation counts were observed. Analyses of geographical variation revealed clear country and continental differences, potentially reflecting both the status of women and the historical development of the discipline in particular regions. Finally, an examination of keywords suggested broad thematic preferences differing by gender. Although this work overall shows that the field of music psychology is more balanced than some other (STEM) fields, it provides a picture to inspire discussions and initiatives to further promote fairness and equity in academia."

@@

suppress-title-page: false

@@

In recent years, meta-scientific attention has increasingly turned toward music psychology, with a focus on how WEIRD [@jakubowski2024] and transparent [@eerola2023transparency] the discipline is. In this study, we examine in particular an indicator of equity within the field: gender equality in authorship. Authorship visibility is critical for academic career progression, influencing recognition, funding, and advancement opportunities. Disparities in gender distribution may signal structural barriers and disciplinary cultures that constrain the diversity of contributors and hinder optimal knowledge production [@ni2021gendered]. Establishing a clear understanding of gender representation in music psychology is therefore essential for supporting a more inclusive and sustainable academic community. Previous studies across various disciplines—including psychology—have consistently revealed persistent gender inequalities in authorship, though the extent of these disparities varies by field [@son2022scientific; @shah2021gender; @rock2021quantifying; @gonzalez2020women].

Academia suffers various equity issues, impacting productivity, innovation, and job satisfaction in the workplace, which-in turn-hugely impact academic progress. Although there are several aspects, the focus of this paper is gender inequity. Women are more at a disadvantage than men, with evidence suggesting that women are less present in positions of power. For instance, in the UK in 2016-17, 24.6% of professors are women [@bhopal2021competing] when compared to the 44% of all grades in the UK academia [@harris2025gender]. A similar situation holds for Italy [24% female professors, @Filandri03082021], and to some extent for the United States [14%, @spoon2023]. A pay gap between men and women in academia is also significant, (e.g., female-male wage ratio is 0.85 in 2020 in the United Kingdom) [@quadlin2023higher]. In a “publish or perish” culture of science and academia [@kiai2019protect], one crucial aspect to focus on is authorship. The number and quality (e.g., journal reputation) of authorships an individual determines crucial career development such as grant funding and future (permanent) positions.

Research across several scientific disciplines show that women tend to be less represented as authors compared to men [@banks2025women; @son2022scientific]. As the topics, approaches, and methods attract differential interest from men and women, it is important to recognise that fields of scholarship and academic disciplines vary considerably in terms of gender distribution [@huang2020]. For example, a large survey of authors by @gonzalez2020women showed that the hard sciences (N = 119,592) had the lowest prevalence of women (14.8%), whereas in the biological and social sciences (N = 262,122) the proportion was substantially higher (43.3%), but not equal. In psychology, which is perhaps the closest benchmark for the present focus on music psychology, female authors accounted for 45.2% of the sample (N = 90,067). Encouragingly, however, studies looking at various year ranges between 1960 until 2024 show a general trend that the ratio of women to men in authorships is improving [@gonzalez2020women; @ioannidis2023gender; @jemielniak2025gender; @sanchez2024analysis]. Yet, there is evidence that this improvement is plateauing [@jemielniak2025gender]. Further, although women are more likely to be first author, women are still generally unfairly underrepresented as the last author [@bruck2023bibliometric; @gonzalez2020women; @rock2021quantifying; @shah2021gender].

Past studies have explored the gender distribution according to geographical regions and in relationship to country demographics. In one study, men out-numbered women more so in high-income compared to low-income countries [@ioannidis2023gender]. However, there are also some unexpected trends: while female representation is lowest in countries like Latin America, Caribbean [in physics journals, @son2022scientific] and in South Korea, Japan, and Austria [in medical journals, @bruck2023bibliometric], females were most highly represented in countries like India and Africa [@bruck2023bibliometric; @son2022scientific]. Human development indicators and gender inequality indicators have also revealed systematic inequalities in academia. However, while national indicators of gender equity is positively correlated with women’s academic success [@chan2020], this does not seem to be as clearly reflected in authorships. Female scientific output is not only lower in countries ranked in lowest quartile for human development and gender equity, but also in countries very high in gender equality; better equality seemed to occur in high and medium-developed countries [@sugimoto2015relationship]. Another recent paper shows that female authorship rates are most balanced in countries with low to intermediate Human Development Index [@sanchez2024analysis]. Overall, while socioeconomic status of certain countries might be a reason for unbalanced female authorship [@ioannidis2023gender], human development might not fully explain such differences [@bruck2023bibliometric; @sanchez2024analysis; @son2022scientific], and further research is required to better understand the trends.

Another aspect that differs between distribution of gender in authorship is the research methodologies and topics. In terms of methodologies, women are are less likely to publish quantitative and experimental studies, but rather publish qualitative work [@ashmos2011gendering; @nunkoo2020three; @sebo2020gender; @thelwall2019gender; @zhang2023gender], perhaps as they demonstrate better relational skill involved in qualitative studies, like prolonged relationships and emotional ties with research participants [@sebo2020gender]. Interestingly, such stereotypes occur in the opposite direction: when assessing fictive research abstracts-one with quantitative and one with a qualitative design- scientific quality of the qualitative abstract was ranked more trustworthy and accurate when assigned to a female author [@johansson2002gender]. There are also key different when assessing article keywords. In one analysis of article keywords of medical journals, female-authored articles were more associated with keywords more related to healthcare-related themes like patient involvement, insurance, and quality-of-care compared to males [@bruck2023bibliometric]. Similarly, in computational linguistics, woman are more likely to publish on speech, social, and conversational topics than males; the latter are more likely to publish in formal mathematical, syntax, and semantic approaches [@vogel2012he]. These trends could be explained by the idea that female researchers often conduct research on societal progress, while males prefer research aimed at scientific progress [@zhang2021gender]. Overall, research shows that women choose qualitative research on social and societal topics, while males author papers using more quantitative methods on theoretical and analytical topics.

To improve academic equity issues, it is worthwhile focusing on subdiscipline to provide more 'localised' support to improve the equity in field. This was the aim of the current paper to look within our own field of music psychology. Indeed, the past efforts such as @anglada2019visualizing have already provided a current overview the publications in the music psychology field with suggestions of future directions to improve the field. @eerola2023transparency showcases how Open Science practices (e.g., preregistrations, sharing research materials, data, and analysis scripts) are relatively limited, and encourage authors in the field to start implementing such practices. @jakubowski2024 explores participant samples and musical stimuli used across a wide range of experiment to give an idea of how limited or generalizable the field is. In a similar way to these papers, the current paper aims to give a current overview of gender patterns in authorship in the field of music psychology.

@@

Our aim is to find out what is the gender distribution in the specialist journals of music psychology. More specifically, we want to identify what is the proportion of different types of authorships (single, first, coauthors, and last authorships) for men and women in the published papers in the last 25 years? Are there specific patterns in the authorship distributions across countries, topics of the studies, or temporal trends?

@@

Gender attribution was initially based on first names. We recognise that treating gender as a binary category is inherently problematic. Gender is a complex and multidimensional social construct. However, consistent with prior meta-science research on gender and authorship [@son2022scientific; @shah2021gender; @rock2021quantifying; @ni2021gendered; @gonzalez2020women; @wais2006] and that non-binary authorship generally is less than 0.1% [@son2022scientific], we adopt a binary classification—male and female—for analytical purposes and assume that first names allow for a reasonable, though imperfect, attribution of gender. We used the _genderize API_ [@wais2006], which predicts gender from first names and can be supplemented with country information derived from author affiliations to improve accuracy. This method resolved the gender of 89.3% of authors with a probability greater than 0.90. Only `89` names had a low attribution probability (< 0.55). Unattributed cases were then checked manually, resulting in `185` manual corrections. After this process, `32` names remained ambiguous and `27` were unknown—some likely due to data entry errors in Scopus (e.g. only initials or surnames). These `59` cases were excluded from the dataset. It is likely that the due to challenge of attributing gender accurately not all the gender attributions are correct. Although genderize.io has been shown to achieve 96.6% accuracy in a diverse multinational test database without using a country of origin information -- 98% with the information included -- it also underperforms (82% accuracy) in Asian names [@vanhelene2024inferring]. However, the error rates of these processes have been previously been shown to be non-biased, i.e. showing similar number of mistakes for both genders [@sebo2021performance;@vanhelene2024inferring;@thelwall2019gender].

We carried out manual corrections of the gender attributions where we were familiar with the author or the database had the full first name coded with initials, or the second name used as the first name. This led to `53` corrections.

We defined four types of authorship positions: single authorship, first authorship (which does not include papers with a single author), coauthorship, and last authorship. Coauthorship includes all positions other than first and last. These positions carry different academic prestige: first authorship is typically associated with the primary contributor, while last authorship is often held by senior researchers with established reputations [@tscharntke2007author].

For the analyses, we analysed gender disparities in authorship by comparing the frequency of these authorship types between male and female authors. Our analysis focused on the female author proportion (FAP) and utilised odds ratios (ORs) to compare the likelihood of occupying each authorship position by gender.

@@

A summary of the authorship types and positions are given in Table 1. While the single-authored papers are in minority (8.6%) in this sample overall, there does not seem to be a significant difference between men and women authoring such articles (_p_ = .21); 38.05% of single-authored papers are women with an odds ratio of 0.90 [95% 0.78-1.05] suggesting no clear gender difference despite the figures leaning towards higher frequency of male authors. The overall OR for women as first authors was 1.41 [95% CI: 1.28–1.54], indicating that the odds of first authorship are 41% higher for females compared to the male authors (Fisher's exact test _p_ < .001). For coauthorship positions, the OR was 1.00 [0.91–1.09], suggesting no substantial gender difference (_p_ = .96). In contrast, the odds ratio for women as last authors was 0.73 [0.67–0.81], indicating that the odds of last authorship are 27% lower for women compared to men and significantly different (_p_ < .001).

@@

```
The variation of the number of authors across the papers is considerable. Here we have the maximum of `{r} max(s$n)` authors in a paper and a median of `{r} median(s$n)` (M=`{r} round(mean(s$n),2)`, SD=`{r} round(sd(s$n),2)`). Previous research has calculated the collaboration index across gender [@gonzalez2020women], suggesting that male author tend to have larger number of coauthors (M = 1.5) than female (M = 1.34) in psychology. Here the average author numbers for male and female authors (disregarding authorship positions) indicate similar difference, higher mean author numbers for males (M = `{r} round(S$m[2],2)`, [`{r} round(S$LCI[2],2)`-`{r} round(S$UCI[2],2)`]) than for females (M = `{r} round(S$m[1],2)`, [`{r} round(S$LCI[1],2)`-`{r} round(S$UCI[1],2)`]), _t_(4640.58) = 7.21, _p_ < .001, Cohen's _d_ = -0.21, 95% CI
[-0.26, -0.15]. This analysis, however, does not allow us to distinguish the authorship roles in the counts.

To illustrate how the actual number of authors in a paper impacts female first author likelihood, we calculated the odds ratios across the number of authors in paper, pooling the papers with 7 or more authors together (N=`{r} sum(df$author_N>6)`,  `{r} round(sum(df$author_N>6)/nrow(df)*100,1)`%). We added the single-authored papers to the plot to indicate the odds of being female to be single authors compared to all studies with multiple authors. The results are shown in Figure 2. We can observe that while female authors are not more likely to be single authors than men, they tend to be more likely to be first authors than males in papers with 2 or 3 authors; the odds ratios for first authorship position for female authors for papers with 2 authors is 1.55 [130-1.85], and for 3 authors the odds ratio is 1.61 [1.35-1.92], both significantly different from equal proportion. The last authorship positions for female authors are consequently lower, odds ratio of 0.65 [0.54–0.77] for 2 authors and 0.59 [0.49–0.71] for 3 authors, both also significant at _p_ < .001 level. At four and five authors, the higher odds ratio for female is still evident (OR of 1.37 both both), but the last authorship positions are no longer relative rare for papers with 4 and 5 authors (OR 0.90 and 0.95, respectively). At six authors and above, the odds ratios indicate the same trend (higher ratio for first author, lower for last) although due to lower number of observations, these are no longer significantly different.

@@

For comparison, the closest disciplinary benchmark comes from psychology, where @gonzalez2020women analyzed the gender distribution of 74,413 authors publishing in psychology journals in 2009. Women accounted for 45.2% of the sample, indicating near gender parity overall. However, substantial variation was observed across sub-disciplines: developmental psychology (57.0%) and educational psychology (53.8%) showed higher proportions of female authors, whereas biological psychology (39.4%), experimental psychology (37.5%), and mathematical psychology (25.8%) showed lower representation. Across the full dataset, women were slightly overrepresented in first authorship positions (OR = 1.10) but underrepresented in last authorship positions (OR = 0.86). This pattern resembles the trends identified in the present study.

@@

```{r}
#| id: citations
#| eval: true
#| echo: false
#| warning: false
#| message: false
source(here("scripts","citations.R"))
options(scipen = 9999)
```
One potential difference established in previous bibliometric analyses of gendered authorship is citations [@west2013role;@chatterjee2021gender]. Here we tested whether the citations -- as indexed by Scopus -- show differences across gender. The median citation for studies with female lead authors is `{r} citestats_first$Md[1]` [`{r} citestats_first$CI_lower[1]`–`{r} citestats_first$CI_upper[1]`], and for males, the numbers are identical (Md=`{r} citestats_first$Md[2]` [`{r} citestats_first$CI_lower[2]`–`{r} citestats_first$CI_upper[2]`]) and the difference is not statistically significant ($\chi^2$(1)=`{r} round(stats_first$statistic,2)`, _p_ = `{r} round(stats_first$p.value,3)`) using rank-based Wilcoxon test. The same comparison of citations for studies with last authors yields similar results `{r} citestats_last$Md[1]` [`{r} citestats_last$CI_lower[1]`–`{r} citestats_last$CI_upper[1]`], and for males, the numbers are similar (Md=`{r} citestats_last$Md[2]` [`{r} citestats_last$CI_lower[2]`–`{r} citestats_last$CI_upper[2]`]). Again, the difference is not statistically significant ($\chi^2$(1)=`{r} round(stats_last$statistic,2)`, _p_ = `{r} round(stats_last$p.value,3)`). If we assume that we can weight in the gender contribution of all authors in a publication to its citation count, we observe a minor difference, where median citations for female authors is `{r} citestats_all$Md[1]` [`{r} citestats_all$CI_lower[1]`–`{r} citestats_all$CI_upper[1]`], and for males, the central measure are statistically significantly higher (Md=`{r} citestats_all$Md[2]` [`{r} citestats_all$CI_lower[2]`–`{r} citestats_all$CI_upper[2]`], $\chi^2$(1)=`{r} round(stats_all$statistic,2)`, _p_ < `{r} round(stats_all$p.value,4)`). For reference, prior study of citation patterns across gender in psychology has reported a small effect favoring men (M = 16.57) over women (M = 15.90) [@gonzalez2020women].

@@

We also explored whether the open access status of the articles is associated with gender. Out of `{r} as.numeric(sum(t))` articles, `{r} round(as.numeric(t[1]/sum(t))*100,2)`% are Open Access in this sample, as indexed by Scopus. For first-authored articles, female odds ratio for Open Access is `{r} round(author_OA$Odds_ratio[1],2)` [`{r} round(author_OA$CI_low[1],2)`–`{r} round(author_OA$CI_high[1],2)`], suggesting nearly 50% higher odds associated with publishing open access by women as compared to men. If we observe only the last authorship status of the publications, the difference vanishes with the odds ratio indisguishable from even division (1.00 and the confidence interval), OR = `{r} round(author_OA$Odds_ratio[3],2)` [`{r} round(author_OA$CI_low[3],2)`–`{r} round(author_OA$CI_high[3],2)`].

@@

Past studies have identified consistent geographical patterns in female authorship. For example, a large-scale analysis of psychology publications found that 46.5% of authorships were by women, with European countries showing a lower average proportion of 42.8% [@gonzalez2020women]. To examine geographical differences more closely, we calculated the odds of female authorship by the countries that have published at least 60 publications in the sample, as shown in Figure 3.

```{r}
#| id: countrywise
#| eval: true
#| echo: false
#| warning: false
#| message: false
#| output: asis
#| fig-width: 9
#| fig-height: 8
#| fig.cap: > 
#|   Female author odds ratio across the prolific countries publishing music psychology (with more than sixty publications).

@@

Study data, analysis scripts and supporting information is available at GitHub, <https://tuomaseerola.github.io/gender_in_music_psych>.

>>>>

abstract: "Academia suffers from various equity issues. One such issue is gender inequality in authorship, where many studies show that women are less represented compared to men. Due to authorship differences across disciplines, the goal of this study is to provide a current overview of the gender distribution of authorship within music psychology. A total of 3,373 papers published between 2000 and 2025 across five core music psychology journals were analysed with respect to author gender and authorship role (first, last, solo, and coauthor). In addition, we explored gender patterns related to citations, open-access publishing, and keywords. Overall, women authors were in the minority, accounting for 40.2% of all authors. However, the distribution of authorship roles diverged notably from this overall proportion: women were more likely to be first authors (odds ratio, OR = 1.41), equally likely to be solo authors (OR = 1.11) or coauthors (OR = 1.00), but less likely to be last authors (OR = 0.73). These patterns have become more pronounced over the past 15 years. No substantial gender differences in citation counts were observed. Analyses of geographical variation revealed clear country and continental differences, potentially reflecting both the status of women and the historical development of the discipline in particular regions. Finally, an examination of keywords suggested broad thematic preferences differing by gender. Although this work overall shows that the field of music psychology is more balanced than some other (STEM) fields, it provides a picture to inspire discussions and initiatives to further promote fairness and equity in academia."

@@

suppress-title-page: true

@@

In recent years, meta-scientific research has increasingly been taken up by music psychology, examining how Western, Educated, Industrialized, Rich and Democratic  [WEIRD, @jakubowski2024] and transparent [@eerola2023transparency] the discipline is. In the current article, we examine a particular indicator of equity within the field: gender equality in authorship. Authorship visibility is critical for academic career progression, as it influences recognition, funding, and advancement opportunities. Disparities in gender distribution may signal structural barriers and disciplinary cultures that constrain the diversity of contributors and hinder optimal knowledge production [@ni2021gendered]. Previous studies across various disciplines—including psychology—have consistently revealed persistent gender inequalities in authorship, though the extent of these disparities varies by field [@son2022scientific; @shah2021gender; @rock2021quantifying; @gonzalez2020women]. To support a more inclusive and sustainable academic community, it is essential to establish a better understanding of gender representation in more localised subfield; in this case, in the field of music psychology.

Academia across all disciplines suffers various equity issues, impacting productivity, innovation, and job satisfaction in the workplace, which—in turn—hugely impact academic progress, leadership opportunities, and visibility throughout a career [@allen2021work]. Although there are several issues, this paper focuses on one particular issue: gender inequity. Although we recognise that gender is a multidimensional social construct, for the purposes of the current paper, and because non-binary authorship generally is less than 0.1% [@son2022scientific], we remain consistent and comparable gender and authorship research [@gonzalez2020women;@ni2021gendered;@rock2021quantifying;@shah2021gender;@son2022scientific;@wais2006] and generally adopt a binary (woman-man) classification. In disciplines overall, women are more at a disadvantage than men, with evidence suggesting that women are less present in positions of power. For instance, in the UK in 2016-17, 24.6% of professors were women [@bhopal2021competing] when compared to the 44% of all grades in the UK academia [@harris2025gender]. A similar situation holds for Italy [24% female professors, @Filandri03082021], and to greater extent for the United States [14%, @spoon2023]. A substantial pay gap exists between men and women in academia; for example, a woman-man wage ratio was 0.85 in 2020 in the United Kingdom [@quadlin2023higher]. In a “publish or perish” culture of science and academia [@kiai2019protect], one crucial aspect to focus on is authorship. The number and quality (e.g., journal reputation) of authorships an individual has garnered can determine crucial career development such as grant funding and future (permanent) positions.

Research across several scientific disciplines show that women tend to be less represented as authors compared to men [@banks2025women; @son2022scientific]. As the topics, approaches, and methods attract differential interest from men and women, it is important to recognise that fields of scholarship and academic disciplines vary considerably in terms of gender distribution [@huang2020]. When investigating the gender composition of publications, a large survey of authors by @gonzalez2020women showed that the hard sciences (N = 119,592) had the lowest prevalence of women (14.8%), whereas in the biological and social sciences (N = 262,122) the proportion of woman authors was substantially higher (43.3%), but not equal. In psychology, which is perhaps the closest benchmark for the present focus on music psychology, woman authors accounted for 45.2% of the sample (N = 90,067). Encouragingly, studies looking at various year ranges between 1960 until 2024 show a general trend that the ratio of women to men in authorships is improving [@gonzalez2020women; @ioannidis2023gender; @jemielniak2025gender; @sanchez2024analysis]. However, there is evidence that this improvement is plateauing [@jemielniak2025gender]. Further, there are differences in terms of authorship position. Authorship position is important as it typically reflects career positions: primary contribution is typically associated with the primary contributor, which often are graduate students or postdocs, while last authorship is often held by senior researchers with established reputations [@tscharntke2007author]. Although women are more likely to be the first author, women are still generally unfairly underrepresented as the last author [@bruck2023bibliometric; @gonzalez2020women; @rock2021quantifying; @shah2021gender].

HERE

Past studies have explored the gender distribution according to geographical regions and in relationship to country demographics. In one study, men out-numbered women more in high-income compared to low-income countries [@ioannidis2023gender]. However, there are also some unexpected trends: while female representation is lowest in Latin America, Caribbean [in physics journals, @son2022scientific] and in South Korea, Japan, and Austria [in medical journals, @bruck2023bibliometric], women were most highly represented in India and Africa [@bruck2023bibliometric; @son2022scientific].

Interestingly, while national indicators of gender equity are positively correlated with women’s academic success [@chan2020], this does not seem to be as clearly reflected in authorships. Scientific output from women is not only lower in countries ranked in the lowest quartile for human development and gender equity, but better equality seemed to occur in high and medium-developed countries [@sugimoto2015relationship]. Another recent paper shows that authorship rates between men and women are most balanced in countries with low to intermediate Human Development Index [@sanchez2024analysis]. Overall, while the socioeconomic status of certain countries might be a reason for unbalanced female authorship [@ioannidis2023gender], human development might not fully explain such differences [@bruck2023bibliometric; @sanchez2024analysis; @son2022scientific], and further research is required to better understand the trends.

Another aspect that differs between distribution of gender in authorship is the research methodologies. Women are less likely to publish quantitative and experimental studies than qualitative work [@ashmos2011gendering; @nunkoo2020three; @sebo2020gender; @thelwall2019gender; @zhang2023gender], perhaps as they demonstrate the better relational skill involved in qualitative studies, like prolonged relationships and emotional ties with research participants [@sebo2020gender]. When assessing fictive research abstracts—one with quantitative and one with a qualitative design—scientific quality of the qualitative abstract was ranked more trustworthy and accurate when assigned to a female author [@johansson2002gender], showing further evidence that of female stereotypes.

Research also shows there that topics may vary by gender as shown by article keyword usage. In one analysis of article keywords of medical journals, articles authored by women were predominantly associated with keywords on healthcare-related themes like patient involvement, insurance, and quality-of-care, while articles authored by men were associated with keywords associated with higher publication citations, such as ‘phase II-III trials’, ‘anti-body based therapy’ males [@bruck2023bibliometric]. Similarly, in computational linguistics, women are more likely to publish on speech, social, and conversational topics than men; men are more likely to publish on formal mathematical, syntax, and semantic approaches [@vogel2012he]. These trends could be explained by evidence that research by women often conduct research on societal progress, while research by men is aimed more at scientific progress [@zhang2021gender]. Overall, research shows that women tend to produce qualitative research on social and societal topics, while men use more quantitative methods on theoretical and analytical topics.

@@

To improve academic equity issues, it is worthwhile focusing on subdiscipline to provide more ‘localised’ support to improve the equity in field. Correspondingly, the aim of the current paper was to look within our own field of music psychology. This builds on past efforts in the music psychology field that have provided a current overview as well as future directions and improvements for: a) publications [@anglada2019visualizing], b) participant samples and stimuli for better generalization of findings [@jakubowski2024], and c) current practices and implementation of Open Science practices [e.g., preregistrations, sharing research materials, data, and analysis scripts, @eerola2023transparency]. Our aim was to determine gender distribution in the specialist journals of music psychology. More specifically, we want to identify the proportion of different types of authorships (single, first, coauthors, and last authorships) for men and women in the published papers in the last 25 years and to explore specific patterns in the authorship distributions across countries, topics of the studies, or temporal trends. By giving a current overview of gender patterns in authorship in the field of music psychology, we hope this stimulates further discussion of identifying and implementing methods to combat such inequities for future generations.

@@

Gender attribution was initially based on given names. As mentioned in the introduction, we recognise that treating gender as a binary category is inherently problematic. Gender is a complex and multidimensional social construct. However, consistent with prior meta-science research on gender and authorship [@son2022scientific; @shah2021gender; @rock2021quantifying; @ni2021gendered; @gonzalez2020women; @wais2006] and the fact that non-binary authorship generally is less than 0.1% [@son2022scientific], although this is not necessarily the true picture of the non-binary authorship situation, we nevertheless adopt a binary classification—male and female—for analytical purposes and assume that given names allow for a reasonable, though imperfect, attribution of gender. We used the _genderize API_ [@wais2006], which predicts gender from given names and can be supplemented with country information derived from author affiliations to improve accuracy. This method resolved the gender of 89.3% of authors with a probability greater than 0.90. Only 89 names had a low attribution probability (< 0.55). Unattributed cases were then checked manually, resulting in 185 manual corrections. After this process, 32 names remained ambiguous and 27 were unknown—some likely due to data entry in Scopus such as erroneous switch between given names and surnames. Some errors likely also stemmed from initials for given names, although we recognise that certain authors might prefer to use initials for publications to avoid clear gender bias. Nonetheless, these 59 cases of these errors were excluded from the dataset. It is likely that the due to challenge of attributing gender accurately not all the gender attributions are correct. Although genderize.io has been shown to achieve 96.6% accuracy in a diverse multinational test database without using a country of origin information -- 98% with the information included -- it also underperforms (82% accuracy) in Asian names [@vanhelene2024inferring]. However, the error rates of these processes have been previously been shown to be non-biased, i.e. showing similar number of mistakes for both genders [@sebo2021performance;@vanhelene2024inferring;@thelwall2019gender].

We carried out manual corrections of the gender attributions where we were familiar with the author or the database had the full given name coded with initials, or the second name used as the given name. This led to 53 corrections. For the purpose of the current study, we inferred authors based on outward gender presentation, though we acknowledge that for future studies, which are not database driven, a more accurate attribution would be to explicitly ask confirmation of such inferences.

We defined four types of authorship positions: single authorship, first authorship (which does not include papers with a single author), coauthorship, and last authorship. Coauthorship includes all positions other than first and last. These positions carry different academic prestige: first authorship is typically associated with the primary contributor, while last authorship is often held by senior researchers with established reputations [@tscharntke2007author]. We note that there might be some ambiguity for two-author papers, where the second author could be either someone who contributed less significantly than the publication or a more senior researcher. For the scope of the current paper, as it is difficult from names alone to attribute career seniority and it is more common for last authors to be senior, we assume that last position is the more senior position.

For the analyses, we analysed gender disparities in authorship by comparing the frequency of these authorship types between man and woman authors. Our analysis focused on the proportion of woman authors and utilised odds ratios (ORs) to compare the likelihood of occupying each authorship position by gender.

@@

A summary of the authorship types and positions are given in Table 1. While the single-authored papers are in minority (8.6%) in this sample, there does not seem to be a significant difference between men and women authoring such articles (_p_ = .21); 38.05% of single-authored papers are authored by women with an odds ratio of 0.90 [95% 0.78-1.05] suggesting no clear gender difference despite the figures leaning towards higher frequency of male authors. The overall OR for women as first authors was 1.41 [95% CI: 1.28–1.54], indicating that the odds of first authorship are 41% higher for females compared to the male authors (Fisher's exact test _p_ < .001). For coauthorship positions, the OR was 1.00 [0.91–1.09], suggesting no substantial gender difference (_p_ = .96). In contrast, the odds ratio for women being as the last authors was 0.73 [0.67–0.81], indicating that the odds of last authorship are 27% lower for women compared to men, which was a significant difference (_p_ < .001).

@@

```
The variation of the number of authors across the papers is considerable. Here we have the maximum of `{r} max(s$n)` authors in a paper and a median of `{r} median(s$n)` (M=`{r} round(mean(s$n),2)`, SD=`{r} round(sd(s$n),2)`). Previous research has calculated the collaboration index across gender [@gonzalez2020women], suggesting that male authors tend to have larger number of coauthors (M = 1.5) than female (M = 1.34) in psychology. Here the average author numbers for male and female authors (disregarding authorship positions) indicate similar difference, higher mean author numbers for males (M = `{r} round(S$m[2],2)`, [`{r} round(S$LCI[2],2)`-`{r} round(S$UCI[2],2)`]) than for females (M = `{r} round(S$m[1],2)`, [`{r} round(S$LCI[1],2)`-`{r} round(S$UCI[1],2)`]), _t_(4640.58) = 7.21, _p_ < .001, Cohen's _d_ = -0.21, 95% CI
[-0.26, -0.15]. This analysis, however, does not allow us to distinguish the authorship roles in the counts.

To illustrate how the actual number of authors in a paper impacts female first author likelihood, we calculated the odds ratios across the number of authors in paper, pooling the papers with 7 or more authors together (N=`{r} sum(df$author_N>6)`,  `{r} round(sum(df$author_N>6)/nrow(df)*100,1)`%). We added the single-authored papers to the plot to indicate the odds of being female to be single authors compared to all studies with multiple authors. The results are shown in Figure 2. We can observe that while women authors are not more likely to be single authors than men, they tend to be more likely to be first authors than men in papers with 2 or 3 authors; the odds ratios for first authorship position for women authors for papers with 2 authors is 1.55 [130-1.85], and for 3 authors the odds ratio is 1.61 [1.35-1.92], both significantly different from equal proportion. The last authorship positions for women authors are consequently lower, odds ratio of 0.65 [0.54–0.77] for 2 authors and 0.59 [0.49–0.71] for 3 authors, both also significant at _p_ < .001 level. At four and five authors, the higher odds ratio for women is still evident (OR of 1.37 both both), but the last authorship positions are no longer rare for papers with 4 and 5 authors (OR 0.90 and 0.95, respectively). At six authors and above, the odds ratios indicate the same trend (higher ratio for first author, lower for last) although due to lower number of observations, these are no longer significantly different.

@@

For comparison, the closest disciplinary benchmark comes from psychology, where @gonzalez2020women analysed the gender distribution of 74,413 authors publishing in psychology journals in 2009. Women accounted for 45.2% of the sample, indicating near gender parity overall. However, substantial variation was observed across sub-disciplines: developmental psychology (57.0%) and educational psychology (53.8%) showed higher proportions of female authors, whereas biological psychology (39.4%), experimental psychology (37.5%), and mathematical psychology (25.8%) showed lower representation. Across the full dataset, women were slightly overrepresented in first authorship positions (OR = 1.10) but underrepresented in last authorship positions (OR = 0.86). This pattern resembles the trends identified in the present study.

To explore whether men and women authors transition different degrees to last authorship during their career, we analysed authors with at least four publications over a time span of at least seven years (213 in total). We compared the most common authorship position during the first and last halves of their careers. During the early part of the career, the odds ratio for women being last authors was 0.64 [0.34, 1.21] and during the later part 0.68 [0.39, 1.18]. Comparing the early and late halves of the careers for men and women, there was a trend for a difference in the transition, but it was not statistically significant ($\chi^2$(1)=3.36, _p_ = 0.067).

@@

```{r}
#| id: citations
#| eval: true
#| echo: false
#| warning: false
#| message: false
source(here("scripts","citations.R"))
options(scipen = 9999)
```
One potential difference established in previous bibliometric analyses of gendered authorship is citations [@west2013role;@chatterjee2021gender]. Here we tested whether the citations -- as indexed by Scopus -- show differences across gender. The median citation for studies with women lead authors is `{r} citestats_first$Md[1]` [`{r} citestats_first$CI_lower[1]`–`{r} citestats_first$CI_upper[1]`], and for men, the numbers are identical (Md=`{r} citestats_first$Md[2]` [`{r} citestats_first$CI_lower[2]`–`{r} citestats_first$CI_upper[2]`]) and the difference is not statistically significant ($\chi^2$(1)=`{r} round(stats_first$statistic,2)`, _p_ = `{r} round(stats_first$p.value,3)`) using rank-based Wilcoxon test. The same comparison of citations for studies with last authors yields similar results `{r} citestats_last$Md[1]` [`{r} citestats_last$CI_lower[1]`–`{r} citestats_last$CI_upper[1]`], and for men, the numbers are similar (Md=`{r} citestats_last$Md[2]` [`{r} citestats_last$CI_lower[2]`–`{r} citestats_last$CI_upper[2]`]). Again, the difference is not statistically significant ($\chi^2$(1)=`{r} round(stats_last$statistic,2)`, _p_ = `{r} round(stats_last$p.value,3)`). If we assume that we can weight in the gender contribution of all authors in a publication to its citation count, we observe a minor difference, where median citations for women authors is `{r} citestats_all$Md[1]` [`{r} citestats_all$CI_lower[1]`–`{r} citestats_all$CI_upper[1]`], and for men, the central measure are statistically significantly higher (Md=`{r} citestats_all$Md[2]` [`{r} citestats_all$CI_lower[2]`–`{r} citestats_all$CI_upper[2]`], $\chi^2$(1)=`{r} round(stats_all$statistic,2)`, _p_ < `{r} round(stats_all$p.value,4)`). For reference, prior study of citation patterns across gender in psychology has reported a small effect favoring men (M = 16.57) over women (M = 15.90) [@gonzalez2020women].

@@

We also explored whether the open access status of the articles is associated with gender. Out of `{r} as.numeric(sum(t))` articles, `{r} round(as.numeric(t[1]/sum(t))*100,2)`% are Open Access in this sample, as indexed by Scopus. For first-authored articles, odds ratio of Open Access for women is `{r} round(author_OA$Odds_ratio[1],2)` [`{r} round(author_OA$CI_low[1],2)`–`{r} round(author_OA$CI_high[1],2)`], suggesting nearly 50% higher odds associated with publishing open access by women as compared to men. If we observe only the last authorship status of the publications, the difference vanishes with the odds ratio indisguishable from even division (1.00 and the confidence interval), OR = `{r} round(author_OA$Odds_ratio[3],2)` [`{r} round(author_OA$CI_low[3],2)`–`{r} round(author_OA$CI_high[3],2)`].

@@

Past studies have identified consistent geographical patterns in woman authorship. For example, a large-scale analysis of psychology publications found that 46.5% of authorships were by women, with European countries showing a lower average proportion of 42.8% [@gonzalez2020women]. To examine geographical differences more closely, we calculated the odds of woman authorship by the countries that have published at least 60 publications in the sample, as shown in Figure 3.

```{r}
#| id: countrywise
#| eval: true
#| echo: false
#| warning: false
#| message: false
#| output: asis
#| fig-width: 9
#| fig-height: 8
#| fig.cap: > 
#|   Women author odds ratio across the prolific countries publishing music psychology (with more than sixty publications).

@@

Study data, analysis scripts and supporting information is available at GitHub, [https://anonymous.4open.science/status/gender_in_music_psych-C6FE](https://anonymous.4open.science/status/gender_in_music_psych-C6FE).
